---
title: "3 Programming basics"
output:
  html_document:
    df_print: paged
    css: "style.css"
    toc: true
---

[Book](http://rafalab.dfci.harvard.edu/dsbook/r-basics.html)

# Functions in this chapter

[`any`](#any)
[`all`](#all)
[`else`](#else)
[`function`](#function)
[`if`](#if)
[`ifelse`](#ifelse)
[`is.na`](#is-na)
[`map_dbl`](#map_dbl)
[`sapply`](#sapply)
[`search`](#search)

# 3.1 Conditional expressions

<span id="if">`if()`</span>, 
<span id="else">`else`</span>

```{r}
a <- 0

if (a != 0) {
  print(1/a)
} else {
  print("No reciprocal for 0")
}
```

Let’s look at one more example using the US murders data frame:

```{r}
library(dslabs)
data(murders)
murder_rate <- murders$total / murders$population * 10^5
```

Here is a very simple example that tells us which states, if any, have a murder rate lower than 0.5 per 100,000. The if statement protects us from the case in which no state satisfies the condition.

```{r}
ind <- which.min(murder_rate)

if (murder_rate[ind] < 0.5) {
  print(murders$state[ind])
} else {
  print("No state has a murder rate that low")
}
```
```{r}
ind <- which.min(murder_rate)

if (murder_rate[ind] < 0.25) {
  print(murders$state[ind])
} else {
  print("No state has a murder rate that low")
}
```

<span id="ifelse">`ifelse()`</span>

```{r}
a <- c(0, 1, 2, -4, 5)
result <- ifelse(a > 0, 1/a, NA)
result
```
This table helps us see what happened:
```{r, echo=FALSE}
tmp <-
  data.frame(a = a, 
             is_a_positive = a > 0,
             answer1 = 1/a,
             answer2 = NA,
             result = result)
if(knitr::is_html_output()) {
  knitr::kable(tmp, "html") |>
    kableExtra::kable_styling(bootstrap_options = "striped",
                              full_width = FALSE)
} else {
  knitr::kable(tmp, "latex", booktabs = TRUE) |>
    kableExtra::kable_styling(font_size = 8)
}
```

<span class="green">Here is an example of how this function can be readily used to replace all the missing values in a vector with zeros:</span>
<span id="is-na">`is.na()`</span>

```{r}
data("na_example")
no_nas <- ifelse(is.na(na_example), 0, na_example)
sum(is.na(no_nas))
```

Two other functions are <span id="any">`any()`</span> and <span id="all">`all()`</span>. The `any` function takes a vector of logicals and returns `TRUE` if any of the entries are `TRUE`. The `all` function takes a vector of logicals and returns `TRUE` if all of the entries are `TRUE`. Here is an example:

```{r}
z <- c(TRUE, TRUE, FALSE)
any(z)
all(z)
```

# 3.2 Defining functions {#function}

```{r}
avg <- function(x) {
  s <- sum(x)
  n <- length(x)
  s/n
}
```
```{r}
x <- 1:100
identical(mean(x), avg(x))
```

Notice that <span class="green">variables defined inside a function are not saved in the workspace.</span> So while we use `s` and `n` when we call `avg`, the values are created and changed only during the call.

 The general form of a function definition looks like this:
 
 ```
 my_function <- function(VARIABLE_NAME){
  perform operations on VARIABLE_NAME and calculate VALUE
  VALUE
}
 ```
 
 Multiple arguments:
```{r}
avg <- function(x, arithmetic = TRUE) {
  n <- length(x)
  ifelse(arithmetic, sum(x)/n, prod(x)^(1/n))
}
```

# 3.3 Namespaces

```{r}
library(dplyr)
```

Use <span id="search">`search()`</span> to see namespaces in search order

```{r}
search()
```

So what if we want to use the `stats` filter instead of the `dplyr` filter but `dplyr` appears first in the search list? You can force the use of a specific namespace by using double colons (::) like this:

```
stats::filter
```

# 3.4 For-loops

<span id="for">`for()`</span>

```{r}
for (i in 1:5) {
  print(i)
}
```
```{r}
compute_s_n <- function(n){
  x <- 1:n
  sum(x)
}
```

```{r}
m <- 25
s_n <- vector(length = m)
for (n in 1:m) {
  s_n[n] <- compute_s_n(n)
}
```
```{r}
n <- 1:m
plot(n, s_n)
```

# 3.5 Vectorization and functionals

Although for-loops are an important concept to understand, in R we rarely use them. As you learn more R, you will realize that <span class="green">vectorization is preferred over for-loops</span> since it results in shorter and clearer code. We already saw examples in the Vector Arithmetic section. <span class="green>A vectorized function is a function that will apply the same operation on each of the vectors.</span>

```{r}
x <- 1:10
sqrt(x)
```
```{r}
y <- 1:10
x * y
```
To make this calculation, there is no need for for-loops. However, not all functions work this way. For instance, the function we just wrote, compute_s_n, does not work element-wise since it is expecting a scalar. This piece of code does not run the function on each entry of n:

```
n <- 1:25
compute_s_n(n)
```

**Functionals** are functions that help us apply the same function to each entry in a vector, matrix, data frame, or list. Here we cover the functional that operates on numeric, logical, and character vectors: <span id="sapply">`sapply()`</span>.

```{r}
x <- 1:10
sapply(x, sqrt)
```

```{r}
n <- 1:25
s_n <- sapply(n, compute_s_n)
s_n
```
Other functionals are `apply`, `lapply`, `tapply`, `mapply`, `vapply`, and `replicate`. We mostly use sapply, apply, and replicate in this book, but we recommend familiarizing yourselves with the others as they can be very useful.

# 3.6 Exercises

1. What will this conditional expression return?

```{r}
x <- c(1,2,-3,4)

if(all(x>0)){
  print("All Postives")
} else{
  print("Not all positives")
}
```

2. Which of the following expressions is always FALSE when at least one entry of a logical vector x is TRUE?

  a. all(x)
  b. any(x)
  c. any(!x)
  d. all(!x)
  
  d
  
3. The function <span id="nchar">`nchar()`</span> tells you how many characters long a character vector is. Write a line of code that assigns to the object new_names the state abbreviation when the state name is longer than 8 characters.

```{r}
ind <- which(nchar(murders) > 8)
new_names <- ifelse(nchar(murders$state) > 8, murders$abb, murders$state)
new_names
```

4. Create a function `sum_n` that for any given value, say $n$, computes the sum of the integers from 1 to n (inclusive). Use the function to determine the sum of integers from 1 to 5,000.

```{r}
sum_n <- function (n) {
  sum(1:n)
}
sum_n(5000)
```

5. Create a function `altman_plot` that takes two arguments, x and y, and plots the difference against the sum.

```{r}
altman_plot <- function(x, y) {
  plot(x-y, x+y)
}
altman_plot(23,32)
```


6. After running the code below, what is the value of x?

```
x <- 3
my_func <- function(y){
  x <- 5
  y+5
}
```

x = 3

7. Write a function `compute_s_n` that for any given $n$ computes the sum $S_n=1^2+2^2+3^2+\dots+n^2$. Report the value of the sum when $n=10$.

```{r}
compute_s_n <- function(n) {
  total <- 0
  for (x in 1:n) {
    total <- total + x^2
  }
  total
}
```
```{r}
compute_s_n(10)
```


8. Define an empty numerical vector s_n of size 25 using s_n <- vector("numeric", 25) and store in the results of $S_1,S_2,\dots,S_25$ using a for-loop.


```{r}
s_n <- vector("numeric", 25)
compute_sq <- function(n) {
  n^2
}
```
```{r}
for (n in 1:25) {
  s_n[n] <- compute_sq(n)
}
s_n
```

9. Repeat exercise 8, but this time use sapply.

```{r}
s_n <- sapply(1:25, compute_sq)
s_n
```

10. Repeat exercise 8, but this time use <span id="map_dbl">`map_dbl`</span>.


```{r}
# install.packages("purrr")
library("purrr")
```

```{r}
s_n <- map_dbl(1:25, compute_sq)
s_n
```

11. Plot $S_n$ versus $n$. Use points defined by $n=1,\dots,25$.

```{r}
n <- 1:25
plot(n, s_n)
```

12. Confirm that the formula for this sum is $S_n=n(n+1)(2n+1)/6$.

```{r}
compute_s_n(5)
```
```{r}
new_sum <- function(n) {
  n*(n+1)*(2*n+1)/6
}
new_sum(5)
```

```{r}
identical(compute_s_n(5), new_sum(5))
```



