---
title: "3 Programming basics"
output:
  html_document:
    df_print: paged
    css: "style.css"
    toc: true
---

[Book](http://rafalab.dfci.harvard.edu/dsbook/r-basics.html)

# Functions in this chapter

|<a href="#any">`any`</a>|
<a href="#all">`all`</a>|
<a href="#else">`else`</a>|
<a href="#if">`if`</a>|
<a href="#ifelse">`ifelse`</a>|
<a href="#is-na">`is.na`</a>|

# 3.1 Conditional expressions

<span id="if">`if()`</span>, 
<span id="else">`else`</span>

```{r}
a <- 0

if (a != 0) {
  print(1/a)
} else {
  print("No reciprocal for 0")
}
```

Let’s look at one more example using the US murders data frame:

```{r}
library(dslabs)
data(murders)
murder_rate <- murders$total / murders$population * 10^5
```

Here is a very simple example that tells us which states, if any, have a murder rate lower than 0.5 per 100,000. The if statement protects us from the case in which no state satisfies the condition.

```{r}
ind <- which.min(murder_rate)

if (murder_rate[ind] < 0.5) {
  print(murders$state[ind])
} else {
  print("No state has a murder rate that low")
}
```
```{r}
ind <- which.min(murder_rate)

if (murder_rate[ind] < 0.25) {
  print(murders$state[ind])
} else {
  print("No state has a murder rate that low")
}
```

<span id="ifelse">`ifelse()`</span>

```{r}
a <- c(0, 1, 2, -4, 5)
result <- ifelse(a > 0, 1/a, NA)
result
```

<span class="hl3">Here is an example of how this function can be readily used to replace all the missing values in a vector with zeros:</span>
<span id="is-na">`is.na()`</span>

```{r}
data("na_example")
no_nas <- ifelse(is.na(na_example), 0, na_example)
sum(is.na(no_nas))
```

Two other functions are <span id="any">`any()`</span> and <span id="all">`all()`</span>. The `any` function takes a vector of logicals and returns `TRUE` if any of the entries are `TRUE`. The `all` function takes a vector of logicals and returns `TRUE` if all of the entries are `TRUE`. Here is an example:

```{r}
z <- c(TRUE, TRUE, FALSE)
any(z)
all(z)
```






















