---
title: "4 The tidyverse"
output:
  html_document:
    df_print: paged
    css: "style.css"
    toc: true
---

[Book](http://rafalab.dfci.harvard.edu/dsbook/r-basics.html)

# Functions in this chapter

|<a href="filter">`filter`</a>|
<a href="mutate">`mutate`</a>|
<a href="select">`select`</a>|

Up to now we have been manipulating vectors by reordering and subsetting them through indexing. However, once we start more advanced analyses, the preferred unit for data storage is not the vector but the data frame. In this chapter we learn to work directly with data frames, which greatly facilitate the organization of information. We will be using data frames for the majority of this book. We will focus on a specific data format referred to as **tidy** and on specific collection of packages that are particularly helpful for working with tidy data referred to as the **tidyverse**.

```{r}
library(tidyverse)
```

We will learn how to implement the tidyverse approach throughout the book, but before delving into the details, in this chapter we introduce some of the most widely used tidyverse functionality, starting with the **dplyr** package for manipulating data frames and the **purrr** package for working with functions. Note that the tidyverse also includes a graphing package, **ggplot2**, which we introduce later in Chapter 8 in the Data Visualization part of the book; the **readr** package discussed in Chapter 5; and many others. In this chapter, we first introduce the concept of tidy data and then demonstrate how we use the tidyverse to work with data frames in this format.

# 4.1 Tidy data

We say that a data table is in tidy format if each row represents one observation and columns represent the different variables available for each of these observations. The murders dataset is an example of a tidy data frame.

```{r}
library(dslabs)
data(murders)
head(murders)
```

Each row represent a state with each of the five columns providing a different variable related to these states: name, abbreviation, region, population, and total murders.

To see how the same information can be provided in different formats, consider the following example:

```
#>       country year fertility
#> 1     Germany 1960      2.41
#> 2 South Korea 1960      6.16
#> 3     Germany 1961      2.44
#> 4 South Korea 1961      5.99
#> 5     Germany 1962      2.47
#> 6 South Korea 1962      5.79
```

This tidy dataset provides fertility rates for two countries across the years. This is a tidy dataset because each row presents one observation with the three variables being country, year, and fertility rate. However, this dataset originally came in another format and was reshaped for the dslabs package. Originally, the data was in the following format:

```
#>       country 1960 1961 1962
#> 1     Germany 2.41 2.44 2.47
#> 2 South Korea 6.16 5.99 5.79
```
The same information is provided, but there are two important differences in the format: 1) each row includes several observations and 2) one of the variables, year, is stored in the header. For the tidyverse packages to be optimally used, data need to be reshaped into tidy format, which you will learn to do in the Data Wrangling part of the book. Until then, we will use example datasets that are already in tidy format.

Although not immediately obvious, as you go through the book you will start to appreciate the advantages of working in a framework in which functions use tidy formats for both inputs and outputs. You will see how this permits the data analyst to focus on more important aspects of the analysis rather than the format of the data.

# 4.2 Exercises

1. Examine the built-in dataset `co2`. Which of the following is true:

  a. `co2` is tidy data: it has one year for each row.
  b. `co2` is not tidy: we need at least one column with a character vector.
  c. `co2` is not tidy: it is a matrix instead of a data frame.
  d. `co2` is not tidy: to be tidy we would have to wrangle it to have three columns (year, month and value), then each co2 observation would have a row.
  
d

```{r}
# co2
```


2. Examine the built-in dataset `ChickWeight`. Which of the following is true:

  a. `ChickWeight` is not tidy: each chick has more than one row.
  b. `ChickWeight` is tidy: each observation (a weight) is represented by one row. The chick from which this measurement came is one of the variables.
  c. `ChickWeight` is not tidy: we are missing the year column.
  d.` ChickWeight` is tidy: it is stored in a data frame.
  
```{r}
class(ChickWeight)
```

d

3. Examine the built-in dataset `BOD`. Which of the following is true:

  a. `BOD` is not tidy: it only has six rows.
  b. `BOD` is not tidy: the first column is just an index.
  c. `BOD` is tidy: each row is an observation with two values (time and demand)
  d. `BOD` is tidy: all small datasets are tidy by definition.
  
```{r}
BOD
```

c
  
4. Which of the following built-in datasets is tidy (you can pick more than one):

 a. BJsales No
 b. EuStockMarkets Yes
 c. DNase Yes
 d. Formaldehyde Yes
 e. Orange Yes
 f. UCBAdmissions No

# 4.3 Manipulating data frames

## 4.3.1 <span id="mutate">`mutate()`</span>: Add columns

We want all the necessary information for our analysis to be included in the data table. So the first task is to add the murder rates to our murders data frame. The function mutate takes the data frame as a first argument and the name and values of the variable as a second argument using the convention name = values. So, to add murder rates, we use:

```{r}
library(dslabs)
data("murders")
murders <- mutate(murders, 
                  rate = total / population * 10^5)
head(murders)
```

## 4.3.2 Subsetting with <span id="filter">`filter()`</span>

```{r}
filter(murders, rate <= 0.71)
```

## 4.3.3 Selecting columns with <span id="select">`select()`</span>

```{r}
new_table <- select(murders, state, region, rate)
filter(new_table, rate <= 0.71)
```

# 4.4 Exercises

1. Load the `dplyr` package and the `murders` dataset.

```{r}
library(dplyr)
library(dslabs)
data(murders)
```

Use the function mutate to add a murders column named rate with the per 100,000 murder rate as in the example code above. Make sure you redefine murders as done in the example code above ( murders <- [your code]) so we can keep using this variable.

```{r}
murders <- mutate(murders,
                  rate = total / population * 10^5)
head(murders)
```

2. If rank(x) gives you the ranks of x from lowest to highest, rank(-x) gives you the ranks from highest to lowest. Use the function mutate to add a column rank containing the rank, from highest to lowest murder rate. Make sure you redefine murders so we can keep using this variable.

```{r}
murders <- mutate(murders,
       rank = rank(-murders$total))
head(murders)
```

3. With `dplyr`, we can use select to show only certain columns. For example, with this code we would only show the states and population sizes:

`select(murders, state, population) |> head()`

Use `select` to show the state names and abbreviations in `murders`. Do not redefine `murders`, just show the results.

```{r}
select(murders, state, abb) |> head()
```


4. The `dplyr` function `filter` is used to choose specific rows of the data frame to keep. Unlike `select` which is for columns, filter is for rows. For example, you can show just the New York row like this:

`filter(murders, state == "New York")`

You can use other logical vectors to filter rows.

Use filter to show the top 5 states with the highest murder rates. After we add murder rate and rank, do not change the murders dataset, just show the result. Remember that you can filter based on the rank column.

```{r}
filter(murders, rank <= 5.0)
```

5. We can remove rows using the != operator. For example, to remove Florida, we would do this:

`no_florida <- filter(murders, state != "Florida")`
Create a new data frame called no_south that removes states from the South region. How many states are in this category? You can use the function nrow for this.

```{r}
no_south <- filter(murders, region != "South")
nrow(no_south)
```

6. We can also use `%in%` to `filter` with `dplyr`. You can therefore see the data from New York and Texas like this:

`filter(murders, state %in% c("New York", "Texas"))`

Create a new data frame called `murders_nw` with only the states from the Northeast and the West. How many states are in this category?

```{r}
murders_nw <- filter(murders, region %in% c("Northeast", "South"))
nrow(murders_nw)
```

7. Suppose you want to live in the Northeast or West and want the murder rate to be less than 1. We want to see the data for the states satisfying these options. Note that you can use logical operators with `filter`. Here is an example in which we filter to keep only small states in the Northeast region.

`filter(murders, population < 5000000 & region == "Northeast")`

Make sure murders has been defined with rate and rank and still has all states. Create a table called my_states that contains rows for states satisfying both the conditions: it is in the Northeast or West and the murder rate is less than 1. Use select to show only the state name, the rate, and the rank.

```{r}
murders |> 
  filter(rate < 1 & region %in% c("Northeast", "West")) |>
  select(state, rate, rank)
```

# 4.5 The pipe: `|>` or '%>%`

```{r}
murders |> 
  select(state, region, rate) |>
  filter(rate <= 0.71)
```

```{r}
16 |> sqrt() |> log2()
```

# 4.6 Exercises

1. The pipe `|>` can be used to perform operations sequentially without having to define intermediate objects. Start by redefining murder to include rate and rank.

```{r}
murders <- mutate(murders, rate =  total / population * 100000, 
                  rank = rank(-rate))
```
In the solution to the previous exercise, we did the following:

```
my_states <- filter(murders, region %in% c("Northeast", "West") & 
                      rate < 1)

select(my_states, state, rate, rank)
```
The pipe `|>` permits us to perform both operations sequentially without having to define an intermediate variable my_states. We therefore could have mutated and selected in the same line like this:

```
mutate(murders, rate =  total / population * 100000, 
       rank = rank(-rate)) |>
  select(state, rate, rank)
```
Notice that select no longer has a data frame as the first argument. The first argument is assumed to be the result of the operation conducted right before the `|>`.

Repeat the previous exercise, but now instead of creating a new object, show the result and only include the state, rate, and rank columns. Use a pipe |> to do this in just one line.

```{r}
murders |>
  mutate(rate = total / population * 10^5,
         rank = rank(-rate)) |>
  filter(region %in% c("Northeast", "West") &
           rate < 1) |>
  select(state, rate, rank) -> my_states
my_states
```


2. Reset murders to the original table by using data(murders). Use a pipe to create a new data frame called my_states that considers only states in the Northeast or West which have a murder rate lower than 1, and contains only the state, rate and rank columns. The pipe should also have four components separated by three |>. The code should look something like this:

my_states <- murders |>
  mutate SOMETHING |> 
  filter SOMETHING |> 
  select SOMETHING

# 4.7 Summarizing data

An important part of exploratory data analysis is summarizing data. The average and standard deviation are two examples of widely used summary statistics. More informative summaries can often be achieved by first splitting data into groups. In this section, we cover two new dplyr verbs that make these computations easier: summarize and group_by. We learn to access resulting values using the pull function.

## 4.7.1 <span id="summarize">`summarize()`</span>

```{r}
library(dplyr)
library(dslabs)
data(heights)
head(heights)
```

The following code computes the average and standard deviation for females:

```{r}
s <- heights |>
  filter(sex == "Female") |>
  summarise(average = mean(height), standard_deviation = sd(height))
s
```

```{r}
s$average
s$standard_deviation
```

For another example of how we can use the summarize function, let’s compute the average murder rate for the United States. Remember our data table includes total murders and population size for each state and we have already used dplyr to add a murder rate column:

```
murders <- murders |> mutate(rate = total/population*100000)
```

Remember that the US murder rate is **not** the average of the state murder rates:

```{r}
summarise(murders, mean(rate))
```

This is because in the computation above the small states are given the same weight as the large ones. The US murder rate is the total number of murders in the US divided by the total US population. So the correct computation is:

```{r}
us_murder_rate <- murders |>
  summarise(rate = sum(total) / sum(population) * 100000)
us_murder_rate
```

## 4.7.2 Multiple summaries

Suppose we want three summaries from the same variable such as the median, minimum, and maximum heights. The <span id="quantile">`quantile()`</span> function: `quantile(x, c(0.5, 0, 1))` returns the median (50th percentile), the min (0th percentile), and max (100th percentile) of the vector x. We can use it with summarize like this:

```{r}
heights |>
  filter(sex == "Female") |>
  summarise(median_min_max = quantile(height, c(0.5, 0, 1)))
```
However, notice that the summaries are returned in a row each. To obtain the results in different columns, we have to define a function that returns a data frame like this:

```{r}
median_min_max <- function(x) {
  qs <- quantile(x, c(0.5, 0, 1))
  data.frame(median = qs[1],
             minimum = qs[2],
             maximum = qs[3])
}
heights |>
  filter(sex == "Female") |>
  summarise(median_min_max(height))
```

## 4.7.3 Group then summarize with <span id="group_by">`group_by()`</span>

```{r}
heights |> group_by(sex)
```

The result does not look very different from heights, <span class="hl3">except we see Groups: sex [2] when we print the object</span>. Although not immediately obvious from its appearance, <span class="hl3">this is now a special data frame called a grouped data frame,</span> and `dplyr` functions, in particular `summarize`, will behave differently when acting on this object. Conceptually, you can think of this table as many tables, with the same columns but not necessarily the same number of rows, stacked together in one object. When we summarize the data after grouping, this is what happens:

```{r}
heights |>
  group_by(sex) |>
  summarise(average = mean(height), standard_deviation = sd(height))
```
The summarize function applies the summarization to each group separately.

For another example, let’s compute the median, minimum, and maximum murder rate in the four regions of the country using the `median_min_max` defined above:

```{r}
murders |>
  group_by(region) |>
  summarise(median_min_max(rate))
```
















