---
title: "5 Importing data"
output:
  html_document:
    df_print: paged
    css: "style.css"
    toc: true
---

[Book](http://rafalab.dfci.harvard.edu/dsbook/r-basics.html)

# Functions in this chapter

|<a href="#file-create">`file.create`</a>|
<a href="#file-copy">`file.copy`</a>|
<a href="#file-link">`file.link`</a>|
<a href="#file-path">`file.path`</a>|
<a href="#file-remove">`file.remove`</a>|
<a href="#file-rename">`file.rename`</a>|
<a href="#file-symlink">`file.symlink`</a>|
<a href="#list-files">`list.files`</a>|
<a href="#read-csv">`read_csv`</a>|
<a href="#read.csv">`read.csv`</a>|
<a href="#read-lines">`read_lines`</a>|
<a href="#scan">`scan`</a>|
<a href="#spec">`spec`</a>|
<a href="#system-file">`system.file`</a>|
<a href="#view">`View`</a>|


```{r}
filename <- "murders.csv"
dir <- system.file("extdata", package = "dslabs") 
fullpath <- file.path(dir, filename)
file.copy(fullpath, "murders.csv")
```
```{r}
library(tidyverse)
dat <- read_csv(filename)
```

Examine data frame columns <span id="spec">**`spec()`**</span>

```{r}
spec(dat)
```

# 5.1 Paths and the working directory

## 5.1.4 Generating path names

Find the full file names of files in packages etc. <span id="system-file">**`system.file()`**</span>

```{r}
filename <- "murders.csv"
dir <- system.file("extdata", package = "dslabs") 
fullpath <- file.path(dir, filename)
fullpath
```

The function `system.file` provides the full path of the folder containing all the files and directories relevant to the package specified by the package argument. By exploring the directories in dir we find that the extdata contains the file we want:

```{r}
dir <- system.file(package = "dslabs") 
filename %in% list.files(file.path(dir, "extdata"))
```

The `system.file` function permits us to provide a subdirectory as a first argument, so we can obtain the fullpath of the extdata directory like this:

```{r}
dir <- system.file("extdata", package = "dslabs") 
dir
```

Construct Path to File <span id="file-path">**`file.path()`**</span>

```{r}
fullpath <- file.path(dir, filename)
fullpath
```

## 5.1.5 Copying files using paths `file.copy()`

The final line of code we used to copy the file into our home directory used
the function `file.copy`. This function takes two arguments: the file to copy and the name to give it in the new directory.

```{r}
file.copy(fullpath, "murders.csv")
```

<span id="list-files">**`list.files()`**</span> to list files

```{r}
list.files()
```

## File manipulation

- <span id="file-create">**`file.create()`**</span>
creates files with the given names if they do not already exist and truncates them if they do.
```{r}
file.create("test.txt")
```

- <span id="file-exists">**`file.exists()`**</span> returns a logical vector indicating whether the files named by its argument exist.

```{r}
file.exists("test.txt")
```

- <span id="file-rename">**`file.rename()`**</span> 
attempts to rename files (and from and to must be of the same length). Where file permissions allow this will overwrite an existing element of to. 

```{r}
file.rename("test.txt", "test1.txt")
```

- <span id="file-copy">**`file.copy()`**</span>
```
file.copy(from, to, overwrite = recursive, recursive = FALSE,
          copy.mode = TRUE, copy.date = FALSE)
```
```{r}
file.copy("test1.txt", "test2.txt")
```

- <span id="file-append">**`file.append()`**</span>
attempts to append the files named by its second argument to those named by its first.

```{r}
file.append("test2.txt", "test1.txt")
```

- <span id="file-link">**`file.link()`**</span> and <span id="file-symlink">**`file.symlink()`**</span>
make symbolic and hard links on those file systems which support them

```{r}
file.symlink("test1.txt", "testsym.txt")
```

- <span id="file-remove">**`file.remove()`**</span>
attempts to remove the files named in its argument. 

```{r}
file.remove(c("testsym.txt", "test1.txt", "test2.txt"))
```

# 5.2 The readr and readxl packages

In this section we introduce the main tidyverse data importing functions. We will use the murders.csv file provided by the dslabs package as an example. To simplify the illustration we will copy the file to our working directory using the following code:
```
filename <- "murders.csv"
dir <- system.file("extdata", package = "dslabs") 
fullpath <- file.path(dir, filename)
file.copy(fullpath, "murders.csv")
```
## 5.2.1 readr

The readr library includes functions for reading data stored in text file spreadsheets into R. <span class="pink">readr is part of the tidyverse package, or you can load it directly</span>:

```
library(readr)
```

The following functions are available to read-in spreadsheets:


|Function	|Format	|Typical suffix|
|:--|:--|:--|
|read_table	|white space separated values	|txt|
|read_csv	|comma separated values	|csv|
|read_csv2	|semicolon separated values	|csv|
|read_tsv	|tab delimited separated values	|tsv|
|read_delim	|general text file format, must define delimiter	|txt|

Although the suffix usually tells us what type of file it is, there is no guarantee that these always match. We can open the file to take a look or use the function <span id="read-lines">**`read_lines`**</span> to look at a few lines:

```{r}
read_lines("murders.csv", n_max = 3)
```

This also shows that there is a header. Now we are ready to read-in the data into R. From the .csv suffix and the peek at the file, we know to use <span id="read-csv">**`read_csv`**</span>:

```{r}
dat <- read_csv(filename)
```

Note that we receive a message letting us know what data types were used for each column. Also note that dat is a tibble, not just a data frame. This is because read_csv is a tidyverse parser. We can confirm that the data has in fact been read-in with <span id="view">**`View()`**</span>:

```{r}
View(dat)
```

Finally, note that we can also use the full path for the file:

```{r}
dat <- read_csv(fullpath)
```

## 5.2.2 readxl

You can load the readxl package using
```{r}
library(readxl)
```

The package provides functions to read-in Microsoft Excel formats:

|Function	|Format	|Typical suffix|
|:--|:--|:--|
|read_excel	|auto detect the format	|xls, xlsx|
|read_xls	|original format	|xls|
|read_xlsx	|new format	|xlsx|

The Microsoft Excel formats permit you to have more than one spreadsheet in one file. These are referred to as **sheets**. The functions listed above read the first sheet by default, but we can also read the others. <span class="pink">The `excel_sheets` function gives us the names of all the sheets in an Excel file. These names can then be passed to the `sheet` argument in the three functions above to read sheets other than the first.</span>

# 5.3 Exercises

1. Use the `read_csv` function to read each of the files that the following code saves in the files object:
```{r}
path <- system.file("extdata", package = "dslabs")
files <- list.files(path)
files
```

```{r}
testread <- read_csv(file.path(path, files[4]))
```
```{r}
testread <- read_csv(file.path(path, files[7]))
```

2. Note that the last one, the olive file, gives us a warning. This is because the first line of the file is missing the header for the first column.

Read the help file for read_csv to figure out how to read in the file without reading this header. If you skip the header, you should not get this warning. Save the result to an object called dat.

```{r}
dat <- read_csv(file.path(path, files[7]), skip = 1)
```

3. A problem with the previous approach is that we don’t know what the columns represent.

```{r}
names(dat)
```

Use the readLines function to read in just the first line (we later learn how to extract values from the output).

```{r}
readLines(file.path(path, files[7]), n = 1)
```

# 5.4 Downloading files

Another common place for data to reside is on the internet. When these data are in files, we can download them and then import them or even read them directly from the web. For example, we note that because our dslabs package is on GitHub, the file we downloaded with the package has a url:

```{r}
url <- "https://raw.githubusercontent.com/rafalab/dslabs/master/inst/extdata/murders.csv"
dat <- read_csv(url)
```

If you want to have a local copy of the file, you can use the <span id="download-file">**`download.file`**</span> function:

```{r}
download.file(url, "download.csv")
```
This will download the file and save it on your system with the name murders.csv. You can use any name here, not necessarily murders.csv. Note that when using download.file you should be careful as <span class="pink">it will overwrite existing files without warning.</span>

```{r}
file.remove("download.csv")
```

Two functions that are sometimes useful when downloading data from the internet are <span id="tempdir">**`tempdir`**</span> and <span id="tempfile">**`tempfile`**</span>. The first creates a directory with a random name that is very likely to be unique. Similarly, tempfile creates a character string, not a file, that is likely to be a unique filename. So you can run a command like this which erases the temporary file once it imports the data:

```{r}
tmp_filename <- tempfile()
download.file(url, tmp_filename)
dat <- read_csv(tmp_filename)
```
```{r}
file.remove(tmp_filename)
```

# 5.5 R-base importing functions

R-base also provides import functions. These have similar names to those in the tidyverse, for example `read.table`, <span id="read.csv`>read.csv`</span> and `read.delim`. You can obtain an data frame like `dat` using:

```{r}
dat2 <- read.csv(filename)
```


An often useful R-base importing function is <span id="scan">`scan`</span>, as it provides much flexibility. When reading in spreadsheets many things can go wrong. The file might have a multiline header, be missing cells, or it might use an unexpected encoding13. We recommend you read this post about common issues found here: https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/.

With experience you will learn how to deal with different challenges. Carefully reading the help files for the functions discussed here will be useful. With scan you can read-in each cell of a file. <span class="pink">`what` declares the data type to be read in</span>. Here is an example:

```{r}
path <- system.file("extdata", package = "dslabs")
filename <- "murders.csv"
x <- scan(file.path(path, filename), sep = ",", what = "c")
x[1:20]
```
<span class="pink">Note that the tidyverse provides read_lines, a similarly useful function.</span>













