---
title: "6 data.table"
output:
  html_document:
    df_print: paged
    css: "style.css"
    toc: true
---

[Book](http://rafalab.dfci.harvard.edu/dsbook/r-basics.html)

# Functions in this chapter

|[`:=`](#col-eq)|
[`copy`](#copy)|
[`setDT`](#setdt)|

In this book, we use tidyverse packages to illustrate because beginners find the code readable and this permits us to focus on data analysis and statistical concepts. However, there are other approaches to wrangling and analyzing data in R that are faster and better at handling large objects. The data.table package, for example, is one of the most popular and we therefore provide a very quick introduction in this chapter . There are several online resources, including this introduction15 for those wanting to learn more.

```{r}
# install.packages("data.table")
library(data.table)
```

# 6.1 Manipulating data tables

We will provide example code showing the **data.table** approaches to **dplyr**’s `mutate`, `filter`, `select`, `group_by`, and `summarize` shown in Chapter 4. As in that chapter, we will use the 1murders1 dataset:

The first step when using **data.table** is to convert the data frame into a **data.table** object using the <span id="setdt">`setDT`</span> function:

```{r}
library(dslabs)
data(murders)
murders <- setDT(murders)
murders
```

<span class="purple">Without this initial step, most of the approaches shown below will not work.</span>

# 6.1.1 Selecting

Selecting with **data.table** is done in a similar way to subsetting matrices. While with **dplyr** we write:
```
select(murders, state, region)
```
In **data.table**, we use notation similar to what is used with matrices:

```{r}
murders[, c("state", "region")] |> head()
```

## 6.1.2 Adding a column or changing columns

We learned to use the **dplyr** `mutate` function with this example:
```
murders <- mutate(murders, rate = total / population * 100000)
```
**data.table** uses an approach that avoids a new assignment (update by reference). This can help with large datasets that take up most of your computer’s memory. The **data.table** <span id="col-eq">`:=`</span> function permits us to do this:

```{r}
murders[, rate := total / population * 100000]
```

This adds a new column, `rate`, to the table. Notice that, as in **dplyr**, we used `total` and `population` without quotes.

We can see that the new column is added:

```{r}
head(murders)
```

To define new multiple columns, we can use the := function with multiple arguments:

```{r}
murders[, ":="(rate = total / population * 10^5, rank = rank(population))]
head(murders)
```

## 6.1.3 Technical detail: reference versus copy

The **data.table** package is designed to avoid wasting memory. So if you make a copy of a table, like this:
```
x <- data.table(a = 1)
y <- x
```
`y` is actually referencing `x`, it is not an new object: it’s just another name for `x`. Until you change `y`, a new object will not be made. However, the `:= ` function changes by reference so if you change `x`, a new object is not made and `y` continues to be just another name for `x`:
```
x[,a := 2]
y
#>    a
#> 1: 2
```
You can also change `x` like this:
```
y[,a := 1]
x
#>    a
#> 1: 1
```
To avoid this, you can use the <span id="copy">`copy`</span> function which forces the creation of an actual copy:
```
x <- data.table(a = 1)
y <- copy(x)
x[,a := 2]
y
#>    a
#> 1: 1
```
## 6.1.4 Subsetting

With **dplyr**, we filtered like this:
```
filter(murders, rate <= 0.7)
```
With **data.table**, we again use an approach similar to subsetting matrices, except **data.table** knows that `rate` refers to a column name and not an object in the R environment:

```{r}
murders[rate <= 0.7]
```

<span class="purple">Notice that we can combine the filter and select into one succint command.</span> Here are the state names and rates for those with rates below 0.7.

```{r}
murders[rate <= 0.7, .(state, rate)]
```

Compare to the dplyr approach:
```{r}
murders |> dplyr::filter(rate <= 0.7) |> dplyr::select(state, rate)
```

# 6.2 Exercises

1. Load the data.table package and the murders dataset and convert it to data.table object:

```{r}
library(data.table)
library(dslabs)
data(murders)
murders <- setDT(murders)
```

Remember you can add columns like this:
```
murders[, population_in_millions := population / 10^6]
```
Add a murders column named rate with the per 100,000 murder rate as in the example code above.

```{r}
murders[, rate := total / population * 10^5]
murders
```

2. Add a column rank containing the rank, from highest to lowest murder rate.

```{r}
murders[, rank := rank(-rate)]
murders
```

3. If we want to only show the states and population sizes, we can use:

```
murders[, .(state, population)] 
```
Show the state names and abbreviations in murders.
```{r}
murders[, .(state, abb)]
```

4. You can show just the New York row like this:

```
murders[state == "New York"]
```
You can use other logical vectors to filter rows.

Show the top 5 states with the highest murder rates. After we add murder rate and rank, do not change the murders dataset, just show the result. Remember that you can filter based on the rank column.

```{r}
murders[rank <= 5]
```

5. We can remove rows using the != operator. For example, to remove Florida, we would do this:
```
no_florida <- murders[state != "Florida"]
```
Create a new data frame called no_south that removes states from the South region. How many states are in this category? You can use the function nrow for this.

```{r}
no_south <- murders[region != "South"]
nrow(no_south)
```

6. We can also use %in% to filter. You can therefore see the data from New York and Texas as follows:
```
murders[state %in% c("New York", "Texas")]
```
Create a new data frame called murders_nw with only the states from the Northeast and the West. How many states are in this category?

```{r}
murders_nw <- murders[region %in% c("Northeast", "West")]
nrow(murders_nw)
```

7. Suppose you want to live in the Northeast or West and want the murder rate to be less than 1. We want to see the data for the states satisfying these options. Note that you can use logical operators with filter. Here is an example in which we filter to keep only small states in the Northeast region.
```
murders[population < 5000000 & region == "Northeast"]
```
Make sure murders has been defined with rate and rank and still has all states. Create a table called my_states that contains rows for states satisfying both the conditions: they are in the Northeast or West and the murder rate is less than 1. Show only the state name, the rate, and the rank.

```{r}
my_state <- murders[region == c("Northeast", "West") & rate < 1, .(state, rate, rank)]
my_state
```

















